Wrangling coral and macroalgae data to be joinable

### 2024-08-30 julianna_algalMetric_wrangling.Rmd
Updated wrangling code to create CSVs with hard cover estimates + richness/Shannon normalized to cover

# metadata matrix
algae %>% 
  # bring in sand data
  full_join(sand) %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg]), # percent cover of algae
         algae_cover_hard = case_when(
           sand == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (algae_cover/(100-sand))*100 
           )) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, algae_cover, 
         algae_cover_hard)  -> sppMeta
```

@@

```{r join}
algDiv %>% 
  full_join(betaDHabYr) %>% 
  # follow a cleaner naming convention to match Noam's
  rename(algae_richness = Richness,
         algae_shannon = Shannon) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            algae_cover_mean = mean(algae_cover),
            algae_cover_se = sd(algae_cover)/sqrt(N),
            algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
            algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(N),
            algae_richness_mean = mean(algae_richness),
            algae_richness_se = sd(algae_richness)/sqrt(N),
            algae_shannon_mean = mean(algae_shannon),
            algae_shannon_se = sd(algae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE)
            ) -> siteData

@@

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(
    algae_richness_mean = mean(algae_richness), 
    algae_richness_se = sd(algae_richness)/sqrt(n()),
    algae_richness_cv = CV(algae_richness),
    algae_shannon_mean = mean(algae_shannon), 
    algae_shannon_se = sd(algae_shannon)/sqrt(n()),
    algae_shannon_cv = CV(algae_shannon),
    # for total algal cover
    algae_cover_mean = mean(algae_cover),
    algae_cover_se = sd(algae_cover)/sqrt(n()),
    algae_cover_cv = CV(algae_cover),
    # and for % of hard substrate
    algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
    algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(n()),
    algae_cover_hard_cv = CV(algae_cover_hard),
    ) -> alltimeCV
```

>>>>

# metadata matrix
algae %>% 
  # bring in sand data
  full_join(sand) %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg]), # percent cover of algae
         algae_cover_hard = case_when(
           sand == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (algae_cover/(100-sand))*100 
           ),
         hard_cover = 100-sand) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, algae_cover, 
         algae_cover_hard, hard_cover, sand)  -> sppMeta
```

@@

```{r join}
algDiv %>% 
  full_join(betaDHabYr) %>% 
  # follow a cleaner naming convention to match Noam's
  rename(algae_richness = Richness,
         algae_shannon = Shannon) %>% 
  # get algal metrics relative to hard
  mutate(algae_richness_hard = algae_richness/hard_cover,
         algae_shannon_hard = algae_shannon/hard_cover) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            algae_cover_mean = mean(algae_cover),
            algae_cover_se = sd(algae_cover)/sqrt(N),
            algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
            algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(N),
            algae_richness_mean = mean(algae_richness),
            algae_richness_se = sd(algae_richness)/sqrt(N),
            algae_shannon_mean = mean(algae_shannon),
            algae_shannon_se = sd(algae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE),
            # add hard metrics
            algae_richness_hard_mean = mean(algae_richness_hard, na.rm = TRUE),
            algae_richness_hard_se = sd(algae_richness_hard, na.rm = TRUE)/sqrt(N),
            algae_shannon_hard_mean = mean(algae_shannon_hard, na.rm = TRUE),
            algae_shannon_se = sd(algae_shannon_hard, na.rm = TRUE)/sqrt(N),
            hard_cover_mean = mean(hard_cover, na.rm = TRUE),
            hard_cover_se = sd(hard_cover, na.rm = TRUE)/sqrt(N)
            ) -> siteData

@@

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(
    algae_richness_mean = mean(algae_richness), 
    algae_richness_se = sd(algae_richness)/sqrt(n()),
    algae_richness_cv = CV(algae_richness),
    algae_shannon_mean = mean(algae_shannon), 
    algae_shannon_se = sd(algae_shannon)/sqrt(n()),
    algae_shannon_cv = CV(algae_shannon),
    # for total algal cover
    algae_cover_mean = mean(algae_cover),
    algae_cover_se = sd(algae_cover)/sqrt(n()),
    algae_cover_cv = CV(algae_cover),
    # and for % of hard substrate
    algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
    algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(n()),
    algae_cover_hard_cv = CV(algae_cover_hard),
    # including diversity
    algae_richness_hard_mean = mean(algae_richness_hard, na.rm = TRUE),
    algae_richness_hard_se = sd(algae_richness_hard, na.rm = TRUE)/sqrt(n()),
    algae_shannon_hard_mean = mean(algae_shannon_hard, na.rm = TRUE),
    algae_shannon_se = sd(algae_shannon_hard, na.rm = TRUE)/sqrt(n()),
    hard_cover_mean = mean(hard_cover, na.rm = TRUE),
    hard_cover_se = sd(hard_cover, na.rm = TRUE)/sqrt(n())
    ) -> alltimeCV
```

### 2024-08-30 julianna_coral_macroalg_wrangling.Rmd
Updated wrangling code to create CSVs with hard cover estimates + richness/Shannon normalized to cover

# metadata matrix
macroalgae %>% 
  # bring in sand data
  full_join(sand_coral) %>% 
  rename(coral_cover = coral,
         sand_cover = sand) %>% 
  mutate(macroalgae_cover = rowSums(macroalgae[startAlg:endAlg]), # percent cover of algae
         macroalgae_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (macroalgae_cover/(100-sand_cover))*100 
           ),
         # also for corals
         coral_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (coral_cover/(100-sand_cover))*100 
           )
         ) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, macroalgae_cover, 
         macroalgae_cover_hard, sand_cover, coral_cover, coral_cover_hard)  -> sppMeta
```

@@

```{r get richness and Shannon diversity}
sppMeta %>% 
  # these functions use the vegan package
  mutate(macroalgae_richness = specnumber(sppMat),
         macroalgae_shannon = diversity(sppMat, "shannon")) -> macroAlgDiv
```

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            ) -> siteData

@@

            # coral
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            coral_cover_cv = CV(coral_cover),
            coral_cover_hard_cv = CV(coral_cover_hard),

@@

            # macroalgae richness
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            )  -> alltimeCV
```

>>>>

# metadata matrix
macroalgae %>% 
  # bring in sand data
  full_join(sand_coral) %>% 
  rename(coral_cover = coral,
         sand_cover = sand) %>% 
  mutate(hard_cover = 100-sand_cover) %>% 
  mutate(macroalgae_cover = rowSums(macroalgae[startAlg:endAlg]), # percent cover of macroalgae
         macroalgae_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (macroalgae_cover/hard_cover)*100 
           ),
         # also for corals
         coral_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (coral_cover/hard_cover)*100 
           )
         ) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, macroalgae_cover, 
         macroalgae_cover_hard, sand_cover, hard_cover, coral_cover, coral_cover_hard)  -> sppMeta
```

@@

```{r get richness and Shannon diversity}
sppMeta %>% 
  # these functions use the vegan package
  mutate(macroalgae_richness = specnumber(sppMat),
         macroalgae_shannon = diversity(sppMat, "shannon"),
         macroalgae_richness_hard = macroalgae_richness/hard_cover,
         macroalgae_shannon_hard = macroalgae_shannon/hard_cover
         ) -> macroAlgDiv
```

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            hard_cover_mean = mean(hard_cover),
            hard_cover_se = sd(hard_cover)/sqrt(N),

            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            # of hard cover
            macroalgae_richness_hard_mean = mean(macroalgae_richness_hard, na.rm = TRUE),
            macroalgae_richness_hard_se = sd(macroalgae_richness_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_shannon_hard_mean = mean(macroalgae_shannon_hard, na.rm = TRUE),
            macroalgae_shannon_hard_se = sd(macroalgae_shannon_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            ) -> siteData

@@

            # coral
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            coral_cover_cv = CV(coral_cover),
            coral_cover_hard_cv = CV(coral_cover_hard),
            hard_cover_mean = mean(hard_cover),
            hard_cover_se = sd(hard_cover)/sqrt(N),
            hard_cover_cv = CV(hard_cover),

@@

            # macroalgae richness
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),

            # of hard
            macroalgae_richness_hard_mean = mean(macroalgae_richness_hard, na.rm = TRUE),
            macroalgae_richness_hard_se = sd(macroalgae_richness_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_shannon_hard_mean = mean(macroalgae_shannon_hard, na.rm = TRUE),
            macroalgae_shannon_hard_se = sd(macroalgae_shannon_hard, na.rm = TRUE)/sqrt(N),

            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            )  -> alltimeCV
```
